Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
add a note

The panel on developing and supporting software examined the challenges
around scientific software development/support, mainly focused on research
groups that also produce code in various forms. There was widespread agreement
that developing and maintaining software is hard, but best practices can help.
Several stated that documentation is not just for users, and paying attention
to API documentation, tutorials for building and deploying software, along with
documented development practices can be very helpful in bringing  new developers
into a project efficiently.

There was discussion that backward compatibility is not always desirable, and it
can be very costly. This must be balanced with the aims of a given project, and how
many other projects depend on and use the code when backwards incompatible
changes are to be made. There are many examples in the wider open source
software world of strategies for dealing with this, and again best practices
can go a long way to mitigating issues around backwards compatibility. Many
projects live with sub-optimal code for a while, and may allow backwards
compatibility to be broken at agreed upon development points, such as a major
release for a software library.

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as lists where developers do not participate in
mailing lists as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
and Turk's paper on scaling code in the human dimension\todo{add cite}  discuss how development
lists that are welcoming to people actually have many more people contributing.

@@

What about actual software engineering principles, such as modularity and
extensibility? This is how industry maintains software, and ensures it continues
to be useful. Everyone \note{everyone in the audience?  paper authors?  software developers in general?} thinks that you can't rewrite your software, but if it is
modular then you can keep systems up to date. Extensibility will keep you
relevant if it is built into the project. One counterpoint raised by Jason Riedy
was that trying to take advantage of the latest and greatest hardware often
makes this painful, hence the lack of papers mentioning ``GPUs and exotic
hardware.''

The question of whether funders, such as the NSF, can mandate software plans in
much the same way as they do data management plans. Daniel Katz responded that
software is supposed to be described as part of the NSF data management plan,
and that in their definition, data includes software. A comment from Twitter
(@biomickwatson) raised the issue that this requires reviewers and funders who
understand the answers that are given in these plans. Daniel Katz responded
that in programs focused on software or data this can be done effectively, but
agreed that in more general programs this is indeed a problem.

>>>>

The panel on developing and supporting software examined the challenges
around scientific software development and support, mainly focused on research
groups that also produce code in various forms. There was widespread agreement
that developing and maintaining software is hard, but best practices can help.
Several participants stated that documentation is not just for users, and paying attention
to API documentation, tutorials for building and deploying software, along with
documented development practices can be very helpful in bringing  new developers
into a project efficiently.

There was discussion that backward compatibility is not always desirable, and it
can be very costly. This must be balanced with the aims of a given project, and how
many other projects depend on and use the code when backwards incompatible
changes are to be made. There are many examples in the wider open source
software world of strategies for dealing with this, and again best practices
can go a long way to mitigating issues around backwards compatibility. Many
projects live with sub-optimal code for a while, and may allow backwards
compatibility to be broken at agreed-upon development points, such as a major
release for a software library.

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as mailing lists where developers do not participate
as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
and Turk's paper on scaling code in the human dimension\todo{add cite}  discuss how development
lists that are welcoming to people actually have many more people contributing.

@@

What about actual software engineering principles, such as modularity and
extensibility? This is how industry maintains software, and ensures it continues
to be useful. Often, rewriting software is considered to be too costly, but
with a modular design it can be kept up to date. Extensibility is expected to keep
it relevant, if built into the project. One counterpoint raised by Jason Riedy
was that trying to take advantage of the latest and greatest hardware often
makes this painful, hence the lack of papers mentioning ``GPUs and exotic
hardware.''

The question of whether funders, such as the NSF, can mandate software plans in
much the same way as they do data management plans, was raised. Daniel Katz responded that
software is supposed to be described as part of the NSF data management plan,
and that in their \note{who is ``they''? OCI? NSF?} definition, data includes software. A comment from Twitter
(@biomickwatson) raised the issue that this requires reviewers and funders who
understand the answers that are given in these plans. Daniel Katz responded
that in programs focused on software or data this can be done effectively, but
agreed that in more general programs this is indeed a problem.
